Advanced Normalization Tools (ANTS)

### 2009-06-05 Documentation/Tex/ants.tex
updates to pdfs
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@11 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\noindent{\bf Warping and Invertibility.}ANTS uses physical space to define mappings. The origin etc is in the
coordinates of the world in which the picture (e.g. MRI) was
taken. Then, warping between physical coordinates is relatively
easy. Differences in bounding boxes etc present no problem -- assuming
you avoid inconsistent headers i.e. origins, directions, data
orientation. One may use PrintHeader to check the data and run simple,
fast tests (few iterations) to perform sanity checks before running
through loads of data.
An ANTS deformation consists of
a standard naming prefix plus a standard naming extension. We usually 
assume nii but other file types may be used. The standard naming
extension is Warpxvec.nii for the x component of the deformation with
similar naming for y and z. Each component is stored in a separate
scalar image. The value of a voxel of a deformation/warp component
image is the physical space displacement from that voxel. Note that
the inverse mapping is stored in InverseWarpxvec.nii and provides the
mapping in the opposite direction.  Note that an inverse -- in the 
digital domain -- is only approximate as shown in figure~\ref{fig:inv}.
\begin{figure}
\label{fig:inv}
\center \includegraphics[width=0.75\textwidth]{Figures/DeformationsInDigitalDomain.pdf} 
\itkcaption[Digital Invertibility]{Digital invertibility presents some limitations. 
Here, we see that invertibility is not exact but is gained only by interpolation.  
Thus, in three-dimensional scenarios, there are fundamental limits to the 
degree of invertibility that may be achieved.  The second and third voxels -- from left -- 
in image A undergo an expansion by a factor of 2.  That is, under the mapping, 
2 voxels are mapped to 4.  This gives the definition of the Jacobian -- computed 
by  ANTSJacobian -- which is a unitless measure defined by the 
ratio of volumes.  Thus, $\mathcal{J}(\x)=V_1(\phi(\x))/V_2(\x)$ where $V_i$ represents 
the volume in one domain versus the other.  Thus, if $\phi$ -- the mapping -- 
causes expansion, then $\mathcal{J}(\x) > 1$. }
\vspace{-0.1in}
\end{figure}  
A few important notes follow. 
(1) Deformation directionality: Warps/deformations applied to images occur in the opposite direction from warps/deformations applied to points. That is, if we map image B to A using 
\begin{verbatim}
ANTS -m PR[A,B,1,2] -o OUTPUT 
\end{verbatim}
 then we get a deformation called \verb OUTPUTWarp+extensions ~and an affine transform called \verb OUTPUTAffine.txt . 
This composite mapping - when applied to B - will transform B into the space of A. 
However, if we have points defined in B that we want to map to A, 
we have to use \verb OUTPUTInverseWarp ~and the inverse of \verb OUTPUTAffine.txt .  
This is because the domain and range of the map's definition need to be dense in a way that is appropriate for the data type. 
Figure~\ref{fig:inv} illustrates the concept. 
(2) Older Image Formats: older image formats (e.g. Analyze) may not
have proper origin/offset/direction support! In these cases, we
recommend converting to nii and verifying that data overlays properly.
(3) Transorm Composition: Composition of transforms may be achieved with
ComposeMultiTransform.  (4) Warping with inverses and concatenations --
viable when using diffeomorphisms -- are described in
figure~\ref{fig:warp}.
 \begin{figure}
\label{fig:warp}
\includegraphics[width=0.9\textwidth]{Figures/ANTSSyntax2.pdf} 
\itkcaption[ANTS Warping]{Here, we detail how one applies 
a deformation and associated inverses.  We also see how 
WarpImageMultiTransform can be used to concatenate a series 
of transformations.}
\vspace{-0.1in}
\end{figure}

@@

\subsection{ANTS Similarity Terms}
From Gang.

{\em Note: ANTS does not like spaces between brackets in the call to the metrics.}

@@

A second type of multivariate normalization is able to use the T1 and DTI components 
directly in the optimization of the mapping.  An example of this type of call is below, 
where we assume that one has resampled the DT data to the space of T1: 
\begin{verbatim}
ANTS 3 -m PR[T1template.nii,T1subject.nii,1.25,4]
       -m PR[FAtemplate.nii,FAsubject.nii,1,4] 
-o MultiVar -i 30x30x20  -r Gauss[3,0] -t SyN[0.25]
\end{verbatim}
In this case, dense features within white matter are gained by using the DT component  
in addition to the T1 component.   The T1 PR metric, here, is weighted slightly more 
than the FA PR metric, 1.25 vs. 1.0.

>>>>

\noindent{\bf Warping and Invertibility.}  ANTS uses physical space to
define mappings. The origin etc is in the coordinates of the world in
which the picture (e.g. MRI) was taken. Then, warping between physical
coordinates is relatively easy. Differences in bounding boxes etc
present no problem -- assuming you avoid inconsistent headers
i.e. origins, directions, data orientation. One may use PrintHeader to
check the data and run simple, fast tests (few iterations) to perform
sanity checks before running through loads of data.
An ANTS deformation consists of
a standard naming prefix plus a standard naming extension. We usually 
assume nii but other file types may be used. The standard naming
extension is Warpxvec.nii for the x component of the deformation with
similar naming for y and z. Each component is stored in a separate
scalar image. The value of a voxel of a deformation/warp component
image is the physical space displacement from that voxel. Note that
the inverse mapping is stored in InverseWarpxvec.nii and provides the
mapping in the opposite direction.  Note that an inverse -- in the 
digital domain -- is only approximate as shown in figure~\ref{fig:inv}.
\begin{figure}
\label{fig:inv}
\center \includegraphics[width=0.75\textwidth]{Figures/DeformationsInDigitalDomain.pdf} 
\itkcaption[Digital Invertibility]{Digital invertibility presents some limitations. 
Here, we see that invertibility is not exact but is gained only by interpolation.  
Thus, in three-dimensional scenarios in particular, there are fundamental limits to the 
degree of invertibility that may be achieved.  The second and third voxels -- from left -- 
in image A undergo an expansion by a factor of 2.  That is, under the mapping, 
2 voxels are mapped to 4.  This gives the definition of the Jacobian -- computed 
by  ANTSJacobian -- which is a unitless measure defined by the 
ratio of volumes.  Thus, $\mathcal{J}(\x)=V(\phi(\x))/V(\x)$ where $V$ represents 
the volume operation and $\x$, here, may be a small object.  Thus, if $\phi$ -- the mapping -- 
causes expansion, then $\mathcal{J}(\x) > 1$. }
\vspace{-0.1in}
\end{figure}  
A few important notes follow. 
(1) Deformation directionality: Warps/deformations applied to images occur in the opposite direction from warps/deformations applied to points. That is, if we map image B to A using 
\begin{verbatim}
ANTS 2 -m PR[A,B,1,2] -o OUTPUT 
\end{verbatim}
 then we get a deformation called \verb OUTPUTWarp+extensions ~and an affine transform called \verb OUTPUTAffine.txt . 
This composite mapping - when applied to B - will transform B into the space of A. 
However, if we have points defined in B that we want to map to A, 
we have to use \verb OUTPUTInverseWarp ~and the inverse of \verb OUTPUTAffine.txt .  
This is because the domain and range of the map's definition need to be dense in a way that is appropriate for the data type. 
Figure~\ref{fig:inv} illustrates the concept. 
(2) Older Image Formats: older image formats (e.g. Analyze) may not
have proper origin/offset/direction support! In these cases, we
recommend converting to nii and verifying that data overlays properly.
(3) Transorm Composition: Composition of transforms may be achieved with
ComposeMultiTransform.  (4) Warping with inverses and concatenations --
viable when using diffeomorphisms -- are described in
figure~\ref{fig:warp}.
 \begin{figure}
\label{fig:warp}
\includegraphics[width=0.9\textwidth]{Figures/ANTSSyntax2.pdf} 
\itkcaption[ANTS Warping]{Here, we detail how one applies 
a deformation and associated inverses.  We also see how 
WarpImageMultiTransform can be used to concatenate a series 
of transformations.}
\vspace{-0.1in}
\end{figure}

@@

\subsection{ANTS Similarity Terms}
Here is a script that will let you experiment with different similarity term 
and transformation model combinations.  A few are listed, with parameters 
that are a reasonable starting point.
\begin{verbatim}
II=r16slice.nii # change to your own images 
JJ=r64slice.nii 
DIM=2 # two-dimensional images 
ITS=" -i 100x100x30 " # 3 optimization levels 
TSYNWITHTIME=" -t SyN[1,2,0.05]  -r Gauss[3,0.5] "
TGREEDYSYN=" -t SyN[0.25]  -r Gauss[3,0] "
TELAST=" -t Elast[1] -r Gauss[0,3] " 
TEXP="  -t Exp[0.5,10] -r Gauss[0,3] " 
LABELGUIDED=" -m PSE[${II},${JJ},${II},${JJ},0.75,0.1,25,0,10] "
INTMSQ=" -m MSQ[${II},${JJ},1,0.1] "
INTMI=" -m MI[${II},${JJ},1,32] "
INTPR=" -m PR[${II},${JJ},1,4] "
NAME=TEST
INT=$INTPR
TRAN=$TSYN
ANTS $DIM -o $NAME $ITS  $TRAN  $INT
INVW=" -i ${NAME}Affine.txt ${NAME}InverseWarp.nii "
FWDW=" ${NAME}Warp.nii ${NAME}Affine.txt "
WarpImageMultiTransform  2 ${II} a.nii -R ${JJ} $INVW
WarpImageMultiTransform  2 ${JJ} b.nii -R ${II} $FWDW
\end{verbatim}
Convergence occurs under two conditions: (1) either the maximum number of 
iterations are reached at a given optimization level or (2) the slope 
of the change in the optimization objective is negative or very small. 
{\em Note: ANTS does not like spaces between brackets in the call to the metrics.}

@@

A second type of multivariate normalization is able to use the T1 and DTI components 
directly in the optimization of the mapping.  An example of this type of call is below, 
where we assume that one has resampled the DT data to the space of T1: 
\begin{verbatim}
ANTS 3 -m PR[T1template.nii,T1subject.nii,1.25,4]
       -m PR[FAtemplate.nii,FAsubject.nii,1,4] 
-o MultiVar -i 30x30x20  -r Gauss[3,0] -t SyN[0.25]
\end{verbatim}
In this case, dense features within white matter are gained by using the DT component  
in addition to the T1 component.   The T1 PR metric, here, is weighted slightly more 
than the FA PR metric, 1.25 vs. 1.0.  {\em The convergence criterion for 
the multivariate scenario is a slave to the {\bf last} metric you 
pass on the ANTS command line}.
